COFFEE: an Optimizing Compiler\\for Finite Element Local Assembly

### 2014-04-01 paper/bare_conf.tex
a LOT of refactoring

\begin{abstract}
%\boldmath
The finite element method is widely-employed to determine approximated solutions of partial differential equations. Local assembly is one of its fundamental steps, in which a suitable numerical procedure, or kernel, is applied to all elements in the discretized domain of the equation. Since the size of the domain can be huge, a notable problem consists of producing efficient kernels to minimize the cost local assembly.
In this context, we present design, implementation and systematic performance evaluation of COFFEE, a domain-specific compiler for local assembly kernels. COFFEE manipulates kernels' abstract syntax trees by introducing composable optimizations aimed at improving instruction-level parallelism, especially in the form of SIMD vectorization, and register locality. It then generates C code, possibly including AVX or LRBNi intrinsics, which is eventually just-in-time compiled on the underlying architecture and executed. Performance evaluation using a suite of examples of real-world importance showed that speed-ups over non-optimized kernels between 1.32$\times$ and 4.44$\times$ can be achieved.
\end{abstract}

@@

\begin{IEEEkeywords}
Finite element method, compilers, optimizations, simd vectorization 
\end{IEEEkeywords}

@@

\section{Introduction}
In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modelled by means of partial differential equations (PDEs). Numerical techniques, like finite volume method and finite element method, are widely-employed to approximate solutions of PDEs. Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective. The solution is sought in each cell of the discretized domain by applying suitable numerical kernels. As the number of cells can be of the order of millions, a major issue is the time required to execute the computation, which can be hours or days. To address this problem, domain-specific languages (DSLs) have been developed. The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design (based on Finite Volume Method, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to OP2~\cite{pyop2isc}, demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{OP2-hydra}.

@@

Due to the constraints of available compilers and on linear-algebra specialized libraries, we have automated a set of generic and model-driven code transformations in COFFEE\footnote{COFFEE stands for COmpiler For Finite ElemEnt local assembly.}, a compiler for optimizing local assembly kernels. COFFEE is integrated with Firedrake, a system for solving PDEs through the finite element method based on the OP2 abstraction~\cite{firedrake-code}. It supports any problems expressible with this framework. This allows us to evaluate our code transformations in a range of real-world problems, varying key parameters that impact both solution accuracy and kernel cost, namely the polynomial order of the method (from $p=1$ to $p=4$) and the geometry of elements in the discretized domain (2D triangle, 3D tetrahedron, 3D prism).

@@

The general structure of a Firedrake-generated kernel is shown in Figure~\ref{code:general-structure}. Input are a pointer to a 0-valued bi-dimensional array, which will be used to store the stiffness matrix; a pointer to the element's coordinates in the discretized domain; one or more pointers to so called fields, for instance indicating value of velocity or pressure in the element. The output is the evaluated element stiffness matrix. A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse; 2) definition of some constant bi-dimensional arrays, called ``basis functions'' (or derivatives of basis functions) in finite element terminology, initialized to problem-specific double-precision floatin point values; 3) evaluation of the stiffness matrix in an affine loop nest. All elements in the discretized domain share the same basis function arrays, so they are declared as global read-only arrays (i.e. using \texttt{static const} in C language). Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to the various kernel's objects.

@@

\section{Code Transformations}
\label{sec:code-transf}
The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake. One peculiar characteristic is that they all aim at improving the run-time of the \texttt{ijk} loop nest, where the numerical integration takes place. In rare, yet important cases, however, this might not be sufficient to achieve notable speed-ups. Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions. It is worth noting there are circumnstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that its execution time becomes the dominant factor of the overall local assembly run-time. In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure). Given its inherent complexity, a comprehensive study of this problem is left as further work.

@@

% Splitting
\subsection{Expression Splitting}
\label{sec:expr-split}
In complex kernels, like Burgers' in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging. If the number of variables independent of the innermost-loop dimension is close to or greater than the number of available CPU registers, it is likely to obtain poor register reuse. This usually happens when the number of basis function arrays, temporaries introduced by generalized loop-invariant code motion, and problem constants is large. For example, applying loop-invariant code motion to Burgers on a 3D mesh needs 33 temporaries for the \texttt{ijk} loop order, and compiler's hoisting of invariant loads out of the $k$ loop can be inefficient on architectures with a relatively low number of registers. One potential solution to this problem consists of suitably ``splitting'' the computation of the stiffness matrix $A$ into multiple sub-expressions; an example, for the Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}. Splitting an expression has, in general, several drawbacks. Firstly, it increases the number of accesses to $A$ proportially to the ``split factor'', which is the number of sub-expressions produced. Secondly, depending on how the split is executed, it can lead to redundant computation (e.g. the product $det*W3[i]$ is performed times number of sub-expressions in the code of Figure~\ref{code:helmholtz-split}). Finally, it might affect register locality, although this is not the case of the Helmholtz example: for instance, the same array could be accessed in different sub-expressions, requiring a proportional number of loads be performed. Nevertheless, as shown in Section~\ref{sec:perf-results}, the performance gain from improved register reuse along inner dimensions can still be greater, especially if the split factor and the splitting itself use heuristics to minimize the aforementioned issues.

\begin{algorithm}[t]
\small
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~double LI$\_$0[4] ALIGN;\\
~~~~double LI$\_$1[4] ALIGN;\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j])*det*W3[i];\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += LI$\_$1[k]*LI$\_$1[j]*det*W3[i];\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code generated by Firedrake when $split$ is applied to the optimized Helmholtz problem given in Figure~\ref{code:helmholtz-licm}. In this example, the split factor is 2.}
\label{code:helmholtz-split}
\end{algorithm}

@@

\begin{table}[h]
\begin{center}
\begin{tabulary}{1.0\columnwidth}{|C|C|}
\hline
Name (Abbreviation) & Parameter \\\hline\hline
Generalized loop-invariant code motion ($licm$) &   \\ \hline
Padding &  \\ \hline
Data Alignment & \\ \hline
Loop interchange      & loops  \\ \hline
Loop unrolling  & unroll factor \\ \hline
Register tiling & tile size \\ \hline
Outer-product vectorization ($op$-$vect$) & tile size \\ \hline
Assembly splitting ($split$) & split point, split factor \\ \hline
\end{tabulary}
\end{center}
\caption{Overview of code transformations for Firedrake-generated assembly kernels.}
\label{table:code-transformations}
\end{table}

@@

Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations. This high-level specification is translated by the Fenics Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels. ASTs are then passed to COFFEE, which is capable of applying the transformations described in Section~\ref{sec:code-transf}. The output of COFFEE is C code, which is eventually provided to PyOP2~\cite{pyop2isc}, where just-in-time compilation and parallel execution over the unstructured mesh take place. Because of the large number of (possibly parametric) transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem. Autotuning might be used, although at the moment we avoid it to minimize the run-time overhead. Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of the compiler.

The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}. Initially, an AST is inspected, looking for the presence of iteration spaces and domain-specific information provided by the higher layer. If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the non-affine loop over mesh elements is vectorized, can be applied. This feature, currently under development, has been proved to be useful in several Finite-Volume-based applications~\cite{inter-kernel-vect}. The second transformation step is applied if the backend is a manycore machine, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{pyop2isc}. Then, an ordered sequence of optimization steps are executed. Application of $licm$ must precede padding and data alignment, due to the introduction of temporary arrays. Based on a cost model, loop interchange, $split$, and $op$-$vect$ may be introduced. Their implementation is based on analysis and trasformation of the AST. When $op$-$vect$ is selected, the compiler outputs proper AVX (or LRBNi) intrinsics code. Any possible corner cases are handled: for example, if $op$-$vect$ is to be applied, but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.

@@

The cost model is shown in Figure~\ref{algo:applyCostModel}. It takes into account the number of available logical vector registers ($n\_regs$) and the number of variables iterating along the $j$ and $k$ dimensions ($n\_consts$ for independent variables, $n\_outer\_arrays$ for $j$ variables, and $n\_inner\_arrays$ for $k$ variables, assuming the \texttt{ijk} loop order) to estimate unroll-and-jam and split factors when, respectively, $op$-$vect$ and $split$ are used. The $n\_consts$ parameter includes temporary registers to carry out computations, so setting its value is partly driven by heuristics. If a factor is 0, then the corresponding transformation is not applied. The $split$ transformation is triggered whenever the number of hoistable terms is larger than the available registers along the outer dimension (lines 3-8), which is approximated as half of the total (line 2). A split factor of $n$ means that the assembly expression should be ``cut'' into $n$ sub-expressions. Depending on the structure of the assembly expression, each sub-expression might end up accessing a different number of arrays; the cost model is simplified by assuming that all sub-expressions are of the same size. The unroll-and-jam factor for the $op$-$vect$ transformation is determined as a function of the available registers, i.e. those not used for storing hoisted terms (line 9-11). Finally, the profitability of loop interchange is evaluated (line 12-16)

@@

These problems were studied varying both the shape of mesh elements and the polynomial order $p$ of the method. Intuitively, the bigger the element shape and $p$, the larger is the iteration space. Triangles (2D), tetrahedron (3D), and prisms (3D) were tested as element shape. For instance, in the case of Helmholtz with $p=1$, the size of the $j$ and $k$ loops for the three kind of elements is, respectively, $3$, $4$, and $6$. Moving from 2D to 3D has the effect of increasing the number of basis function arrays, since, conceptually, the behaviour of the equation has now to be approximated also along a third axis. On the other hand, the polynomial order affects only the problem size (the three loops $i$, $j$, and $k$, and, as a consequence, the size of $X$ and $Y$ arrays). A range of polynomial orders, from $p=1$ to $p=4$, were tested; higher polynomial orders are excluded from the study because of current Firedrake limitations. In such a large space of problems, the size of the stiffness matrix rarely excees 30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and $p=4$.

@@

\subsection{Impact of Vector-register Tiling}
Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} show the speed-ups achieved by applying $op$-$vect$ on top of $licm$-$ap$ to the Helmoltz and Diffusion kernels, respectively. For each problem instance we report two bars: one indicates the best run-time achieved by autotuning the unroll/unroll-and-jam factors, whereas the other shows the result retrieved by the COFFEE's cost model. In general, there is no substantial difference between the two. This is chiefly because assembly kernels fit the L1 cache, so estimating the run-time based on the quality of register allocation is relatively simple. The rationale behind these results is that, for smaller configurations, vector-register tiling is rarely helpful. This is due to the fact that in the considered problems the number of accessed arrays along the innermost loop dimension is rather small (between 2 and 4), so extensive unrolling is often enough to maximize register re-use when the loops are relatively short. On the other hand, as soon as the iteration space becomes too big, vector-register tiles are essential to minimize pipeline stalls. A maximum speed-up of 1.4$\times$ over $licm$-$ap$ was observed in the Diffusion kernel when using a prismatic mesh and $p=4$. Application of $op$-$vect$ to the Burgers problem induces meaningful slow downs, because of the large number of arrays that, generally, need to be tiled. Expression splitting can be used in combination to $op$-$vect$ to alleviate this issue; this is discussed in Section~\ref{sec:perf-results-split}.

@@

% why split useless in helmholtz/diffusion
% why vect vect+split useless in burgers

@@

@@

COFFEE currently uses a simple cost model and heuristics to steer the optimization process. Many code generators, like those based on the Polyhedral model~\cite{PLUTO} and those driven by domain-knowledge~\cite{modeldriven}, make use of cost models. The other extreme consists of relying on auto-tuning to select the best implementation for a given problem on a certain platform. Notable examples include tuning of small matrix-matrix multiplies in nek5000~\cite{nek5000}, the ATLAS library~\cite{ATLAS}, and FFTW~\cite{FFTW} for fast fourier transforms. In both cases, pruning the implementation space, as we did in Section~\ref{sec:pyop2-compiler}, is fundamental to mitigate complexity and overhead.

@@

This research is partly funded by the MAPDES project and by the Department of Computing at Imperial College London. The authors would like to thank Prof. J. Ramanujam for his invaluable suggestions, and Dr. Lawrence Mitchell and Dr. Francis Russell for their contribution to the PyOP2 project.

>>>>

%The finite element method is widely-employed to approximate solutions of partial differential equations. Local assembly is one of its steps, in which a problem-specific kernel is applied to all elements in the discretized domain of the equation. Since the domain size can be huge, executing efficient kernels is fundamental. Automating their optimization is, however, a challenging issue. Even if an affine loop nest can be generally distinguished, small problem size and complexity of mathematical expressions make it hard to determine a single or unique sequence of successful transformations. In this context, we present COFFEE, a domain-specific compiler for local assembly kernels. COFFEE manipulates kernels’ abstract syntax trees by introducing composable optimizations aimed at improving instruction-level parallelism, especially SIMD vectorization, and register locality. It then generates C code including vector intrinsics. Systematic performance evaluation using a suite of real-world examples shows that speed-ups over non-optimized kernels up to 4.44$\times$ are achievable.

\begin{abstract}
%\boldmath
The finite element method is widely-employed to approximate solutions of partial differential equations. Local assembly is one of its steps, in which a problem-specific kernel is applied to all elements in the discretized domain of the equation. Since the domain size can be huge, executing efficient kernels is fundamental. Automating their optimization is, however, a challenging issue. Even if an affine loop nest can be generally distinguished, small problem size and complexity of mathematical expressions make it hard to determine a single or unique sequence of successful transformations. In this context, we present design and systematic evaluation of COFFEE, a domain-specific compiler for local assembly kernels. COFFEE manipulates abstract syntax trees by introducing composable optimizations aimed at improving instruction-level parallelism, especially SIMD vectorization, and register locality. It then generates C code including vector intrinsics. Performance evaluation using a suite of real-world examples shows that signifcant speed-ups up are achievable.
\end{abstract}

@@

\begin{IEEEkeywords}
Finite element integration, local assembly, compilers, optimizations, simd vectorization 
\end{IEEEkeywords}

@@

\section{Introduction}
In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modelled by means of partial differential equations (PDEs). Numerical techniques, like finite volume method and finite element method, are widely-employed to approximate solutions of PDEs. Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective. The solution is sought in each cell of the discretized domain by applying suitable numerical kernels. As the number of cells can be of the order of millions, a major issue is the time required to execute the computation, which can be hours or days. To address this problem, domain-specific languages (DSLs) have been developed. The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design (based on Finite Volume Method, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to OP2~\cite{pyop2isc}, demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{IstvanHydra}.

@@

Due to the constraints of available compilers and on linear-algebra specialized libraries, we have automated a set of generic and model-driven code transformations in COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, a compiler for optimizing local assembly kernels. COFFEE is integrated with Firedrake, a system for solving PDEs through the finite element method based on the OP2 abstraction~\cite{firedrake-code}. It supports any problems expressible with this framework. This allows us to evaluate our code transformations in a range of real-world problems, varying key parameters that impact both solution accuracy and kernel cost, namely the polynomial order of the method (from $p=1$ to $p=4$) and the geometry of elements in the discretized domain (2D triangle, 3D tetrahedron, 3D prism).

@@

The general structure of a Firedrake-generated kernel is shown in Figure~\ref{code:general-structure}. Input are a pointer to a 0-valued bi-dimensional array, which will be used to store the stiffness matrix; a pointer to the element's coordinates in the discretized domain; one or more pointers to so called fields, for instance indicating value of velocity or pressure in the element. The output is the evaluated element stiffness matrix. A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse; 2) definition of some constant bi-dimensional arrays, called ``basis functions'' (or derivatives of basis functions) in finite element terminology, initialized to problem-specific double-precision floating point values; 3) evaluation of the stiffness matrix in an affine loop nest. All elements in the discretized domain share the same basis function arrays, so they are declared as global read-only arrays (i.e. using \texttt{static const} in C language). Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to the various kernel's objects.

@@

\section{Code Transformations}
\label{sec:code-transf}
The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake. One peculiar characteristic is that they all aim at improving the run-time of the \texttt{ijk} loop nest, where the numerical integration takes place. In rare, yet important cases, however, this might not be sufficient to achieve notable speed-ups. Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions. It is worth noting there are circumstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that its execution time becomes the dominant factor of the overall local assembly run-time. In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure). Given its inherent complexity, a comprehensive study of this problem is left as further work.

In a local assembly kernel, although the bulk of the computation is carried out by the \texttt{ijk} affine loop nest, the typically small problem size and the complexity of the stiffness matrix evaluation make it hard to determine a single or unique sequence of successful transformations. Even though the memory access pattern is fixed, the fact that both data- and iteration-space sizes vary, possibly substantially, among different problems, render the optimization process challenging. Relatively to machine peak, while some problems achieve very good performance when applying a specific set of transformations, there is also an ample space of cases for which a more elaborated optimization strategy is required. For example, the Burgers problem in Figure~\ref{code:burgers}, given the large number of arrays accessed, will suffer from high register pressure, whereas the Helmholtz problem in Figure~\ref{code:helmholtz} will not; this intuitively suggests that the two problems require a different treatment, based on an in-depth analysis of data and iteration spaces. Furthermore, domain-knowledge enables transformations that a general-purpose compiler could not apply, making the optimization space even larger. In this section, we provide a non-exhaustive list of transformations that local assembly kernels could benefit from. All of them are automated in COFFEE.

@@

% Splitting
\subsection{Expression Splitting}
\label{sec:expr-split}
In complex kernels, like Burgers' in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging. If the number of variables independent of the innermost-loop dimension is close to or greater than the number of available CPU registers, it is likely to obtain poor register reuse. This usually happens when the number of basis function arrays, temporaries introduced by generalized loop-invariant code motion, and problem constants is large. For example, applying loop-invariant code motion to Burgers on a 3D mesh needs 33 temporaries for the \texttt{ijk} loop order, and compiler's hoisting of invariant loads out of the $k$ loop can be inefficient on architectures with a relatively low number of registers. One potential solution to this problem consists of suitably ``splitting'' the computation of the stiffness matrix $A$ into multiple sub-expressions; an example, for the Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}. Splitting an expression has, in general, several drawbacks. Firstly, it increases the number of accesses to $A$ proportionally to the ``split factor'', which is the number of sub-expressions produced. Secondly, depending on how the split is executed, it can lead to redundant computation (e.g. the product $det*W3[i]$ is performed times number of sub-expressions in the code of Figure~\ref{code:helmholtz-split}). Finally, it might affect register locality, although this is not the case of the Helmholtz example: for instance, the same array could be accessed in different sub-expressions, requiring a proportional number of loads be performed. Nevertheless, as shown in Section~\ref{sec:perf-results}, the performance gain from improved register reuse along inner dimensions can still be greater, especially if the split factor and the splitting itself use heuristics to minimize the aforementioned issues.

\begin{algorithm}[t]
\small
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
%~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
~~~~// Same code as in Figure~\ref{code:helmholtz-licm} up to line 15 \\
%~~// Declaration of basis function matrices \\
%~~\\
%~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
%~~~~double LI$\_$0[4] ALIGN;\\
%~~~~double LI$\_$1[4] ALIGN;\\
%~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
%~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
%~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
%~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j])*det*W3[i];\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += LI$\_$1[k]*LI$\_$1[j]*det*W3[i];\\
$\rbrace$
\caption{Local assembly code generated by Firedrake when $split$ is applied to the optimized Helmholtz problem given in Figure~\ref{code:helmholtz-licm}. In this example, the split factor is 2.}
\label{code:helmholtz-split}
\end{algorithm}

@@

\begin{table}[h]
\begin{center}
\begin{tabulary}{1.0\columnwidth}{|C|C|}
\hline
Name (Abbreviation) & Parameter \\\hline\hline
Generalized loop-invariant code motion ($licm$) &   \\ \hline
Padding &  \\ \hline
Data Alignment & \\ \hline
Loop interchange      & loops  \\ \hline
Loop unrolling  & unroll factor \\ \hline
Outer-product vectorization ($op$-$vect$) & tile size \\ \hline
Expression splitting ($split$) & split point, split factor \\ \hline
\end{tabulary}
\end{center}
\caption{Overview of code transformations for Firedrake-generated assembly kernels.}
\label{table:code-transformations}
\end{table}

@@

Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations. This high-level specification is translated by the Fenics Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels. ASTs are then passed to COFFEE, which is capable of applying the transformations described in Section~\ref{sec:code-transf}. The output of COFFEE is C code, which is eventually provided to PyOP2~\cite{pyop2isc}, where just-in-time compilation and parallel execution over the unstructured mesh take place. Because of the large number of (possibly parametric) transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem. Auto-tuning might be used, although at the moment we avoid it to minimize the run-time overhead. Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of the compiler.

The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}. Initially, an AST is inspected, looking for the presence of iteration spaces and domain-specific information provided by the higher layer. If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the non-affine loop over mesh elements is vectorized, can be applied. This feature, currently under development, has been proved to be useful in several Finite-Volume-based applications~\cite{inter-kernel-vect}. The second transformation step is applied if the backend is a manycore machine, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{pyop2isc}. Then, an ordered sequence of optimization steps are executed. Application of $licm$ must precede padding and data alignment, due to the introduction of temporary arrays. Based on a cost model, loop interchange, $split$, and $op$-$vect$ may be introduced. Their implementation is based on analysis and transformation of the AST. When $op$-$vect$ is selected, the compiler outputs proper AVX (or LRBNi) intrinsics code. Any possible corner cases are handled: for example, if $op$-$vect$ is to be applied, but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.

@@

The cost model is shown in Figure~\ref{algo:applyCostModel}. It takes into account the number of available logical vector registers, $n\_regs$, and the number of unique variables accessed: $n\_consts$ counts variables independent of both $j$ and $k$ loops, $n\_outer\_arrays$ counts $j$-dependent variables, and $n\_inner\_arrays$ counts $k$-dependent variables, assuming the \texttt{ijk} loop order. These are used to estimate unroll-and-jam and split factors when, respectively, $op$-$vect$ and $split$ are used. The $n\_consts$ parameter includes temporary registers to carry out computations, so setting its value is partly driven by heuristics. If a factor is 0, then the corresponding transformation is not applied. The $split$ transformation is triggered whenever the number of hoistable terms is larger than the available registers along the outer dimension (lines 3-8), which is approximated as half of the total (line 2). A split factor of $n$ means that the assembly expression should be ``cut'' into $n$ sub-expressions. Depending on the structure of the assembly expression, each sub-expression might end up accessing a different number of arrays; the cost model is simplified by assuming that all sub-expressions are of the same size. The unroll-and-jam factor for the $op$-$vect$ transformation is determined as a function of the available registers, i.e. those not used for storing hoisted terms (line 9-11). Finally, the profitability of loop interchange is evaluated (line 12-16)

@@

These problems were studied varying both the shape of mesh elements and the polynomial order $p$ of the method. Intuitively, the bigger the element shape and $p$, the larger is the iteration space. Triangles (2D), tetrahedron (3D), and prisms (3D) were tested as element shape. For instance, in the case of Helmholtz with $p=1$, the size of the $j$ and $k$ loops for the three kind of elements is, respectively, $3$, $4$, and $6$. Moving from 2D to 3D has the effect of increasing the number of basis function arrays, since, conceptually, the behaviour of the equation has now to be approximated also along a third axis. On the other hand, the polynomial order affects only the problem size (the three loops $i$, $j$, and $k$, and, as a consequence, the size of $X$ and $Y$ arrays). A range of polynomial orders, from $p=1$ to $p=4$, were tested; higher polynomial orders are excluded from the study because of current Firedrake limitations. In such a large space of problems, the size of the stiffness matrix rarely exceeds 30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and $p=4$.

@@

\subsection{Impact of Vector-register Tiling}
\label{sec:perf-eval-opvect}
Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} show the speed-ups achieved by applying $op$-$vect$ on top of $licm$-$ap$ to the Helmoltz and Diffusion kernels, respectively. For each problem instance we report two bars: one indicates the best run-time achieved by auto-tuning the unroll/unroll-and-jam factors, whereas the other shows the result retrieved by the COFFEE's cost model. In general, there is no substantial difference between the two. This is chiefly because assembly kernels fit the L1 cache, so estimating the run-time based on the quality of register allocation is relatively simple. The rationale behind these results is that, for smaller configurations, vector-register tiling is rarely helpful. This is due to the fact that in the considered problems the number of accessed arrays along the innermost loop dimension is rather small (between 2 and 4), so extensive unrolling is often enough to maximize register re-use when the loops are relatively short. On the other hand, as soon as the iteration space becomes too big, vector-register tiles are essential to minimize pipeline stalls. A maximum speed-up of 1.4$\times$ over $licm$-$ap$ was observed in the Diffusion kernel when using a prismatic mesh and $p=4$. Application of $op$-$vect$ to the Burgers problem induces meaningful slow downs, because of the large number of temporary arrays that need to be tiled, which exceeds the number of available logical registers on the underlying machine. Expression splitting can be used in combination to $op$-$vect$ to alleviate this issue; this is discussed in Section~\ref{sec:perf-results-split}.

We have also used the Intel Architecture Code Analyzer (IACA) tool~\cite{IACA} to demonstrate that the improved run-time is actually due to increased register re-use. By applying the tool to the Helmholtz problem on a tetrahedral mesh and $p=4$, it can be noticed that the number of clock cycles to execute one iteration of the $j$ loop decreases by roughly 17$\%$, on the Sandy Bridge, when using $op$-$vect$. This is because of less pressure on both data (cache) ports available in the processor.

@@

The $split$ transformation was also tried in combination with $op$-$vect$ ($split$-$op$-$vect$), although the cost model prevents the application of the latter on the Sandy Bridge. Despite performance improvements up to 1.22$\times$, $split$-$op$-$vect$ never outperforms $split$. This is because of two reasons: for small split factors, like 1 and 2, the data space to be tiled is still too big, and register spilling still affects the run-time; for higher ones, sub-expressions become so small that, as explained in Section~\ref{sec:perf-eval-opvect}, full or partial unrolling of the loop nest allows achieving a certain degree of register re-use.

@@

\subsection{Comparison with Fenics Form Compiler's built-in Optimizations}
We have modified the Fenics Form Compiler (FFC) to make it return an abstract syntax tree representation of a local assembly kernel, rather than plain C code, so as to enable code transformations in COFFEE. Besides Firedrake, FFC is used in the FEniCS project~\cite{Fenics}. In FEniCS, FFC can apply its own model-driven optimizations to local assembly kernels~\cite{quadrature1}, which mainly consist of loop-invariant code motion and elimination, at code generation time, of floating point operations involving zero-valued columns in basis function arrays. The FEniCS' loop-invariant code motion is, however, definitely different from COFFEE's, since it is based on expansion of arithmetic operations to identify terms that are invariant of the whole loop nest. Thus, depending on the way expansion is performed, operation count does not necessarily decrease. On the other hand, elimination of zero-valued columns, which are due to certain properties of the input finite element problem, has the effect of introducing indirection arrays in the generated code. This kind of optimization is currently under development in COFFEE, although it will differ from FEniCS' by avoiding non-contiguous memory accesses, which affect vectorization because of the need for gather operations, at the price of removing only some of the zero-columns.

Table~\ref{table:comparison-to-FFC-opt} summarizes the results of the comparison between FEniCS' (FFC) optimizations and COFFEE on the Sandy Bridge. Slow downs in Burgers occur in presence of small iteration spaces (triangular mesh, $p \in [1, 2]$; tetrahedral mesh, $p \in [1, 2]$; prismatic mesh, $p = 1$), and generally much better than the maximum one reported in table, obtained with a triangular mesh and $p = 1$. This is a result of zero-columns removal in FEniCS: indirection arrays prevent auto-vectorization, which significantly impacts performance as soon as the stiffness matrix becomes bigger than 12$\times$12. However, with the forthcoming zero-removal optimization in COFFEE, we expect to outperform FEniCS even in these corner cases as well. In the cases of Helmholtz and Diffusion, the minimum speed-ups are, respectively, 1.10 and 1.18 (2D mesh, $p=1$), and they keep increasing with polynomial order and element shape up to the values illustrated in the table.

\begin{table}[h]
\begin{center}
\begin{tabulary}{1.0\columnwidth}{|C|C|C|}
\hline
Problem & Max slow down & Max speed up \\\hline\hline
Helmholtz & / & 11.72$\times$ \\
Diffusion & / & 14.68$\times$ \\
Burgers & 2.24$\times$ & 1.61$\times$ \\\hline
\end{tabulary}
\end{center}
\label{table:comparison-to-FFC-opt}
\caption{Summary of performance comparison between FEniCS and COFFEE on the Sandy Bridge machine. The slow down and speed up columns compare the best COFFEE implementation and the FEniCS implementation with optimizations enabled.}
\end{table}

@@

COFFEE currently uses a simple cost model and heuristics to steer the optimization process. Many code generators, like those based on the Polyhedral model~\cite{PLUTO} and those driven by domain-knowledge~\cite{modeldriven}, make use of cost models. The opposite extreme consists of relying on auto-tuning to select the best implementation for a given problem on a certain platform. Notable examples include tuning of small matrix-matrix multiplies in nek5000~\cite{nek5000}, the ATLAS library~\cite{ATLAS}, and FFTW~\cite{FFTW} for fast fourier transforms. In both cases, pruning the implementation space, as we did in Section~\ref{sec:pyop2-compiler}, is fundamental to mitigate complexity and overhead.

@@

This research is partly funded by the MAPDES project and by the Department of Computing at Imperial College London. The authors would like to thank Dr. Lawrence Mitchell and Dr. Francis Russell for their invaluable suggestions and their contribution to the Firedrake project.
